% Chapter 2: Background and Related Work

\section{Science Mapping and Scientific Document Representations}

Science mapping turns relations among papers, journals, authors, subject categories, or topics into spatial form. Classical bibliometric maps use citation, co-citation, bibliographic coupling, co-authorship, co-word occurrence, or journal/category assignment to make scientific systems navigable \parencite{borner2003visualizing}. Tools such as VOSviewer arrange objects so that spatial proximity reflects a chosen similarity relation \parencite{vaneck2010vosviewer}, while global overlay maps show how activity occupies predefined disciplinary regions \parencite{leydesdorff2013global}. The important point is that every map is conditional on the relation used to build it.

Bibliometric relations encode scholarly behavior, but they are not identical to semantic proximity. Two papers may be close in citation space because they share references or communities, and two semantically similar papers may not cite one another. Text-based representations offer a complementary route. The vector-space tradition in information retrieval made documents comparable in a common numerical space \parencite{salton1983vector}; later dense-vector models encoded semantic associations in continuous coordinates \parencite{mikolov2013word2vec}. For this thesis, the operational value is that every retained document can be assigned coordinates, so distances, neighborhoods, centroids, dispersion, and movement can be measured.

Scientific text is not generic text. Titles and abstracts contain specialized terminology, compressed research claims, and field-specific conventions. Domain-adapted models such as SciBERT were developed because scientific corpora differ systematically from general language corpora \parencite{beltagy2019scibert}. SPECTER introduced document-level scientific embeddings learned with citation-informed supervision \parencite{cohan2020specter}, and SPECTER2 extends this line within SciRepEval for scientific document representation \parencite{singh2023scirepeval}. SPECTER2 is therefore a suitable representation for comparing scientific works as documents, while remaining one model-dependent representation rather than a neutral map of science.

\section{From Semantic Location to Field Morphology}

Most uses of document embeddings begin with location: where is a paper, topic, or field in semantic space? Location is important, but incomplete. A centroid summarizes the average position of a field while discarding spread, local packing, hub concentration, and spectral structure. The same field can be close to another in semantic location, far from it in morphology, and still share a domain label.

The morphological intuition has roots in work on diversity, cognitive distance, and novelty. Stirling's framework separates variety, balance, and disparity \parencite{stirling2007diversity}; cognitive-distance research treats proximity and distance as conditions for learning and recombination \parencite{nooteboom2007cognitive}; studies of atypical combinations show that scientific novelty can arise from unusual bridges between existing knowledge regions \parencite{uzzi2013atypical}. This thesis adapts that intuition to document embeddings: a field is not only a point or a label, but a distribution whose shape can be summarized.

Embedding-space morphology is the operational form of that idea. Dispersion describes how far papers spread around the field centroid. Local density describes how tightly papers pack into nearest-neighbor neighborhoods. Hubness describes whether a small subset of papers becomes disproportionately central in nearest-neighbor relations, a known issue in high-dimensional spaces \parencite{aggarwal2001surprising,radovanovic2010hubness}. Spectral structure summarizes how many principal directions are needed to describe variation. Temporal evolution then asks how those same structural dimensions change across publication windows. Longitudinal science-mapping and interdisciplinarity studies similarly treat field structure as something that can differentiate, integrate, and change over time \parencite{vanraan2000growth,porter2009interdisciplinary,rafols2010networkcoherence,leydesdorff2011indicators}.

The gap addressed here is therefore narrower than a general critique of science maps. Existing maps answer useful questions about links, topics, classifications, and visual relationships. I use OpenAlex as an open, auditable bibliographic infrastructure \parencite{priem2022openalex}, while treating its subfields as pragmatic analytical units rather than natural boundaries. Large bibliographic databases differ by field, document type, language, and metadata completeness \parencite{martinmartin2021google}, and OpenAlex has documented strengths and limits \parencite{culbert2025reference}. Within those constraints, the thesis asks what becomes visible when field-level paper distributions are measured by shape rather than only by label, location, or projected map.
