% Chapter 2: Background and Related Work

\section{Science Mapping and Scientific Document Representations}

Science mapping is part of the broader quantitative study of how science is organized and how it changes over time \parencite{fortunato2018science}. Its practical aim is to make large scientific systems readable. It does this by turning relations among papers, journals, authors, subject categories, or topics into spatial form. Classical bibliometric maps use citation, co-citation, bibliographic coupling, co-authorship, co-word occurrence, or journal/category assignment to describe connections \parencite{borner2003visualizing}. Tools such as VOSviewer place objects so that spatial proximity reflects a chosen similarity relation \parencite{vaneck2010vosviewer}, and global overlay maps show how activity occupies predefined disciplinary regions \parencite{leydesdorff2013global}. A map therefore depends on the evidence used to construct it.

Different mapping traditions measure different relations. Direct citation records an explicit link from one work to another. Bibliographic coupling links papers that use similar reference bases. Co-citation links older works that are later cited together, so it depends on subsequent recognition. Co-word methods use repeated lexical association. Taxonomies assign papers to predefined or algorithmic categories. These relations can all be useful, but they should not be treated as interchangeable. Boyack and Klavans compared co-citation, bibliographic coupling, direct citation, and a citation-text hybrid, showing that different citation approaches produce different clustering behavior and different accuracy results \parencite{boyack2010citation}.

This difference matters when a map is used as evidence. A co-citation map can emphasize older publications that have become common reference points. Bibliographic coupling is more sensitive to the reference lists of current papers. Co-word maps may react quickly to new terminology, while taxonomies can give stable labels at the cost of forcing papers into a limited classification system. The choice of relation defines what kind of closeness is being measured. For this reason, a map should be read as an operational construction, not as a complete picture of the scientific system.

Citation relations encode scholarly behavior: shared references, intellectual lineage, communities, and recognition. Semantic proximity is another object. Papers can be close in citation space without using similar language, and semantically similar papers may not cite each other. Text-based representations address this side of the problem by assigning documents coordinates in a common numerical space. The vector-space tradition in information retrieval established this operational view of documents \parencite{salton1983vector}, and later dense-vector models encoded semantic associations in continuous coordinates \parencite{mikolov2013word2vec}. Once documents have coordinates, distances, neighborhoods, centroids, dispersion, and movement become measurable.

A document embedding compresses text into a vector. This is richer than a raw count of words because the model is trained to place related texts near one another according to a learning objective. The resulting coordinates still depend on the model, the training data, and the signal used during training. Kozlowski and colleagues make this point clearly for science-of-science data: text-based embeddings and graph-based embeddings represent different aspects of article relatedness, with semantic models focusing on content and relational models focusing more on citation or social structure \parencite{kozlowski2021semantic}. Scientific-domain training matters for the same reason. Titles and abstracts contain specialized terminology, compressed claims, and field-specific writing conventions. Domain-adapted models such as SciBERT respond to the fact that scientific corpora differ from general language corpora \parencite{beltagy2019scibert}. SPECTER introduced document-level scientific embeddings learned with citation-informed supervision \parencite{cohan2020specter}, and SPECTER2 extends that line within SciRepEval for scientific document representation \parencite{singh2023scirepeval}. In this thesis, SPECTER2 supplies a fixed scientific-document space. The representation is model-dependent, yet it supports consistent relative comparisons under one measurement choice.

\section{From Semantic Location to Field Morphology}

Document embeddings are often used to ask where a paper, topic, or field lies in semantic space. A centroid is useful for that purpose because it gives an average position. It also discards most of the distribution: the spread of papers around the center, the tightness of local neighborhoods, the concentration of nearest-neighbor relations, and the number of principal directions needed to describe variation. Publication volume also leaves this structure unresolved. Milojevi{\'c} shows, in a different measurement setting, that the cognitive extent of scientific domains can be studied independently from the number of published papers \parencite{milojevic2015cognitive}. The point for this thesis is similar but not identical: field structure should not be reduced to output size, a label, or a single average location.

The idea that scientific structure has more than one geometric aspect is consistent with work on diversity, cognitive distance, and novelty. Stirling's framework separates variety, balance, and disparity \parencite{stirling2007diversity}; cognitive-distance research treats proximity and distance as conditions for learning and recombination \parencite{nooteboom2007cognitive}; studies of atypical combinations show how novelty can arise from unusual bridges between established knowledge regions \parencite{uzzi2013atypical}. These studies do not define the morphology metrics used here, but they support the same general idea: scientific structure has several dimensions, and a single position or class label is usually too small a description.

Embedding-space morphology turns this distributional view into a measurable profile. It combines four families that describe different features of the same document cloud. Global dispersion measures how far papers spread around the field centroid. Local density describes how tightly papers are grouped in nearest-neighbor neighborhoods. Hubness measures whether a small set of papers appears disproportionately often as neighbors of other papers, a known feature of high-dimensional spaces \parencite{aggarwal2001surprising,radovanovic2010hubness}. Spectral structure describes how variation is distributed across the main directions of the cloud. These properties are kept separate because they are not equivalent: a field can be broad but locally dense, compact but hub-concentrated, or dispersed across many principal directions. Combining them in one profile provides a common basis for comparison without reducing field structure to a single score.

The temporal question also needs a structural view. Scientific organization changes through specialization, emerging topics, new combinations of methods, and changing relations between areas. Longitudinal science-mapping and interdisciplinarity studies provide the broader setting for treating field structure as something that can differentiate, integrate, and change over time \parencite{vanraan2000growth,porter2009interdisciplinary,rafols2010networkcoherence,leydesdorff2011indicators}. Pan and colleagues give a clear example in physics: using PACS codes over 1985--2009, they show increasing interactions among subfields and a growing role for Interdisciplinary Physics in the network core \parencite{pan2012interdisciplinarity}. That type of work studies relations and integration across subfields. The present thesis uses a narrower definition for Research Question 3. Convergence and divergence mean that the distance between fields' structural morphology profiles decreases or increases over time.

The gap addressed here is specific. Existing approaches describe links, topics, labels, positions, and visual neighborhoods. This thesis measures the internal distributional shape of field-level document clouds in a fixed representation space. I use OpenAlex as an open and auditable bibliographic infrastructure \parencite{priem2022openalex}, while treating its subfields as pragmatic analytical units rather than natural boundaries. Large bibliographic databases differ by field, document type, language, and metadata completeness \parencite{martinmartin2021google}, and OpenAlex has documented strengths and limits \parencite{culbert2025reference}. Within those constraints, morphology gives a controlled way to compare how the papers of scientific fields are arranged.
