% Chapter 3: Data and Corpus Construction

\section{Data Source}

This thesis uses OpenAlex as its primary bibliographic data source. OpenAlex is a fully open index of scholarly works, authors, venues, institutions, and research concepts, designed to support large-scale and reproducible analyses of scientific literature \parencite{priem2022openalex}. Its openness is central to the design of this thesis: the corpus construction process can be documented, versioned, audited, and rerun without dependence on proprietary bibliographic infrastructures.

The choice of OpenAlex is also motivated by its disciplinary hierarchy. The empirical analysis requires a unit that is more granular than broad scientific domains but still stable enough to support comparative and temporal measurement. For this reason, the main analytical unit is the OpenAlex \textit{subfield}. Subfields are later aggregated into \textit{fields} and \textit{domains} whenever broader comparisons are required. This hierarchy makes it possible to study scientific morphology at several levels of resolution without redefining the classification system at each stage.

OpenAlex is not treated as a flawless representation of science. Like all large bibliographic databases, it is affected by coverage differences, metadata incompleteness, classification uncertainty, and language bias. Previous comparisons of large bibliographic sources show that open databases can provide broad coverage for science mapping, but also that coverage varies by database, discipline, and document type \parencite{martinmartin2021google}. More recent evidence indicates that OpenAlex reference coverage is broadly comparable to Web of Science and Scopus under cleaned shared-publication comparisons, while also documenting metadata limitations, including lower abstract coverage \parencite{culbert2025reference}. These limitations are especially relevant here because the thesis relies on title and abstract text to construct scientific document embeddings.

The resulting corpus should therefore be understood as a controlled analytical corpus, not as a census of all scientific production. Its purpose is to support comparable semantic measurements across scientific regions, not to reproduce the true publication volume of each discipline.

\section{Corpus Version and Analytical Scope}

The empirical analyses in this thesis are based on a synchronized OpenAlex title--abstract corpus covering publications from 2000 to 2024. The corpus contains 2,378,036 text-eligible OpenAlex works distributed across 252 subfields, 26 fields, and 4 domains. This full corpus is the synchronized thesis snapshot used for data description and validation.

The downstream embedding-based analysis uses a slightly smaller analysis subset. After aligning the corpus with the available SPECTER2 embeddings and applying the requirements for metric computation, the active analysis matrix contains 2,344,927 embedded works across 241 analysis subfields. This distinction is important: the full corpus describes the validated text extraction, whereas the analysis subset defines the observations that enter the embedding-space morphology metrics.

\input{tables/tab_03_corpus_summary}

The difference between the full corpus and the analysis subset is not a contradiction. It reflects the fact that metric computation requires a row-aligned embedding matrix and sufficient observations per analytical unit. The thesis therefore reports both levels: the full text corpus for corpus construction, and the analysis subset for the embedding-space results.

\section{Unit of Analysis}

The primary unit of analysis is the OpenAlex subfield. Each work is retained for a subfield only when its primary-topic subfield matches the planned analytical cell. This primary-topic rule avoids assigning the same paper to multiple subfields in the main corpus and preserves a clear one-work--one-subfield structure for downstream comparison.

The analysis uses three nested levels:

\begin{itemize}
    \item \textbf{Subfield}: the main unit for metric computation and comparison.
    \item \textbf{Field}: an intermediate aggregation level used to summarize related subfields.
    \item \textbf{Domain}: the broadest level, used to compare large scientific regions.
\end{itemize}

This structure is used pragmatically. The thesis does not claim that OpenAlex subfields are natural or definitive boundaries of scientific knowledge. Rather, they provide a transparent and reproducible reference system for comparing the internal semantic morphology of scientific regions. This distinction matters because the thesis studies differences in semantic organization, not the ontological status of disciplinary categories.

\section{Textual Eligibility Criteria}

The corpus is designed specifically for embedding-based analysis. A work is eligible only if it contains enough textual information to support a meaningful document representation. The text used for embedding is the concatenation of title and abstract.

The active extraction applies the following filters.

\input{tables/tab_03_corpus_filters}

Several choices deserve emphasis. First, the corpus is restricted to English-language records. This improves linguistic consistency for the embedding model but introduces a language bias that must be acknowledged when interpreting results. Second, only articles and preprints are retained. This excludes document types whose structure or purpose may differ substantially from research papers. Third, both title and abstract are required, with minimum token thresholds. This prevents very short or uninformative records from entering the semantic space.

The abstract requirement is methodologically necessary but not neutral. Abstract availability varies across fields, publishers, and historical periods. Older records and some disciplines may be disproportionately excluded. For this reason, the corpus construction stage includes explicit validation diagnostics by year, subfield, and domain. The later empirical results should be interpreted as statements about a title--abstract corpus of scientific works, not about all scholarly outputs indexed by OpenAlex.

\section{Sampling Design}

The corpus uses a balanced annual sampling design. For each subfield and publication year, the extraction targets up to 400 eligible works. Formally, for subfield \(s\) and year \(t\), the planned annual sample size is

\[
n_{s,t}^{planned} = \min \left( N_{s,t}^{eligible}, 400 \right),
\]

where \(N_{s,t}^{eligible}\) denotes the number of OpenAlex works satisfying the textual, language, type, and metadata filters in that subfield-year cell.

Since the period covers 25 years, the maximum full-period target per subfield is

\[
400 \times 25 = 10{,}000
\]

works. The complete plan contains 2,431,303 planned works. The synchronized corpus retains 2,378,036 works, implying a total shortfall of 53,267 works. The shortfall is not a validation failure. It reflects the fact that some subfield-year cells contain fewer than 400 eligible works, or that local filtering and backfilling could not fully recover the target.

The design does not redistribute unused capacity from sparse years into denser years. If a subfield-year cell has fewer than 400 eligible works, the missing observations are recorded as shortfall rather than replaced by additional works from other years. This preserves the temporal structure of the corpus and avoids artificially inflating years with stronger metadata coverage.

The purpose of this sampling design is comparability. Without an annual cap, very large and recent fields would dominate the corpus. The balanced design reduces the influence of publication-volume differences and makes it more meaningful to compare semantic dispersion, density, dimensionality, and temporal movement across subfields.

\section{Domain Composition}

The synchronized corpus covers all four OpenAlex domains. Physical Sciences represents the largest share of retained works, followed by Health Sciences, Social Sciences, and Life Sciences.

\input{tables/tab_03_domain_distribution}

The domain distribution reflects both the OpenAlex taxonomy and the availability of eligible title--abstract records under the sampling design. It should not be interpreted as the true distribution of global scientific production. The annual cap and text-eligibility filters deliberately reshape the corpus to support comparative semantic analysis.

\section{Corpus Construction Pipeline}

The corpus construction pipeline is organized as a sequence of reproducible stages. First, the OpenAlex taxonomy is fetched and stored. Second, annual counts are computed for subfields, fields, and domains. Third, a versioned corpus plan and sample plan are constructed. Fourth, sampled works are downloaded from OpenAlex. Fifth, the downloaded records are locally validated against the planned subfield-year cells and eligibility filters. Finally, the validated corpus snapshot is preserved as the processed text corpus and used as input for the embedding pipeline.

The pipeline can be represented schematically as follows:

\begin{center}
\fcolorbox{gray45}{gray97}{%
\begin{minipage}{0.9\textwidth}
\centering
\small
\begin{tabular}{@{}c@{}}
taxonomy $\rightarrow$ annual counts $\rightarrow$ corpus plan $\rightarrow$\\
sample plan $\rightarrow$ downloaded works $\rightarrow$ validated text corpus.
\end{tabular}
\end{minipage}}
\end{center}

The active embedding pipeline then builds the active SPECTER2 embedding set and a row-aligned analysis matrix from this corpus. The analysis matrix used by downstream morphology scripts contains 2,344,927 rows and 241 analysis subfields. This row-aligned matrix is the contract between the corpus, the embedding representation, and the metric computation stages.

The implementation-level artifact names and execution commands are documented in Appendix A.

\section{Validation Diagnostics}

The validation stage checks whether the extracted corpus satisfies the planned design and whether the retained records meet the required metadata and text conditions. The synchronized validation report did not fail. Critical checks are clean: there are no duplicate work identifiers, no rows outside the 2000--2024 period, no rows in 2025 or 2026, no missing subfield, field, or domain identifiers, no missing primary-topic identifiers, no missing title or abstract fields, no records below the minimum title or abstract length thresholds, no non-English rows, and no disallowed work types.

The remaining limitation is coverage rather than validity. Out of 6,300 diagnostic subfield-year rows, 5,493 reach the annual target of 400 works, while 807 remain below target. These below-target cells account for the total shortfall of 53,267 works. Most shortfalls arise because the corresponding subfield-year cells contain fewer than 400 eligible works or because the download and filtering process could not fully recover the planned count.

This validation structure is important for the interpretation of the thesis. The corpus is internally consistent and synchronized with the embedding pipeline, but some parts of the OpenAlex taxonomy are less densely populated under the title--abstract eligibility rules. Temporal comparisons in sparse subfields should therefore be interpreted with greater caution.

\section{Implications for the Empirical Analysis}

The corpus construction strategy determines what the later embedding-space metrics can and cannot mean. The metrics in this thesis should be interpreted as indicators of the semantic morphology of a balanced title--abstract corpus. They are not measures of total field size, research funding, institutional capacity, citation impact, or causal scientific development.

Three implications follow.

First, the corpus is designed for \textbf{cross-disciplinary comparability}. The annual cap prevents large subfields from dominating the analysis only because they produce more papers.

Second, the corpus is designed for \textbf{temporal comparability}. The annual structure allows later analyses to construct five-year windows without changing the underlying extraction logic.

Third, the corpus is designed for \textbf{textual consistency}. Each retained work has sufficient title and abstract information to support SPECTER2-based representation.

These choices make the corpus appropriate for the central objective of the thesis: comparing the shape and evolution of scientific regions in a shared semantic embedding space. They also impose boundaries on interpretation. The results describe model-dependent semantic structures derived from OpenAlex title--abstract records. They should not be read as objective maps of science itself, but as reproducible measurements of how scientific works are distributed under a specific data source, taxonomy, text representation, and sampling design.