% Chapter 10: Discussion

\section{What It Means to Measure the Shape of Science}

This thesis shifts the object of measurement from the content of a field to the geometry of its embedded paper distribution. A field is not treated as a list of topics, a citation community, a productivity unit, or a position in a predefined taxonomy. It is treated as a distribution of title--abstract records in a shared SPECTER2 embedding space. Its morphology is the way that distribution occupies the space: how dispersed it is, how tightly its local neighborhoods are packed, how concentrated those neighborhoods are around hubs, how many spectral directions are needed to summarize its variation, and how its position changes over time.

That move makes the analysis complementary to standard science maps rather than a replacement for them. Topic models ask what texts are about. Citation networks ask how works, authors, venues, or fields are connected. Bibliometric indicators often ask how much is produced, cited, or recognized. Embedding-space morphology asks a narrower geometric question: given a representation of scientific documents, what structural shape does a field's collection of papers have? Two fields may be morphologically similar without sharing vocabulary, institutions, methods, or citation practices. Conversely, two fields may be topically related while having different dispersion, density, or temporal movement.

The framework is therefore conditional by design. The measurements are not properties of science in itself. They are properties of a balanced OpenAlex title--abstract corpus, grouped by OpenAlex primary-topic subfields, embedded with SPECTER2, and measured in the original 768-dimensional representation space. This conditionality is not a weakness to hide; it is what makes the result reproducible. The thesis does not claim to reveal the true map of science. It proposes a way to measure one aspect of scientific organization while keeping the data source, taxonomy, representation model, sampling design, and geometric assumptions visible.

The concept of morphology is useful because it separates several notions that are often compressed into one map. A subfield has a semantic location, because its centroid lies somewhere in embedding space. It has a shape, because its papers spread unevenly around that location. It has local organization, because some papers become dense neighbors or hubs. It has a trajectory, because these properties change across windows. It has taxonomic metadata, because OpenAlex assigns it to a field and domain. The empirical chapters show that these dimensions are related, but not interchangeable.

\section{Static Shape, Temporal Change, and Semantic Location}

The static comparison in Chapter~6 establishes the first layer of the result: broad domains have morphological tendencies, but they do not determine morphology. Physical Sciences are generally more dispersed and spectrally extended. Health Sciences are more locally dense and hub-concentrated. Social Sciences and Life Sciences occupy less extreme average positions on several metrics. Yet those domain averages are blunt summaries. Field-level and subfield-level profiles vary substantially within each domain, and the most atypical subfields are not outliers for one common reason. Some are unusual because of dispersion, others because of novelty, local density, hubness, or spectral structure. Static morphology is therefore multidimensional; it cannot be reduced to one axis such as broad versus compact, modern versus traditional, or pure versus applied.

The temporal analysis in Chapter~7 adds a second layer. Across the population of subfields, several metrics move toward contraction or densification: median centroid distance, median kNN distance, and PCA \(D_{80}\) decline from the first to the last window. At the same time, local distance variability can rise, and hubness does not move uniformly. This means that temporal evolution is not a simple story of fields becoming smaller, denser, or more homogeneous. A subfield may become locally compact while becoming more uneven; another may reduce spectral dimensionality while losing hub concentration; another may move far in centroid space without matching the structural trajectory of other fast-moving subfields. Dynamic morphology is a bundle of coordinated and uncoordinated changes.

Chapter~8 turns morphology into a relational object. Field-level profile distances show that taxonomy captures part of the structure, but many nearest morphological neighbors cross domain boundaries. The domain-pair matrix shows lower within-domain distances on average, yet the nearest-neighbor results prevent the binary within/cross distinction from becoming too strong. Over time, the average field-pair distance increases, so the profile space becomes more separated rather than globally convergent. This average divergence coexists with selective convergence: some field pairs move closer through shared changes in local distance and spectral dimensionality, while many others move apart. Morphological similarity is therefore relational but not causal. It identifies structural resemblance, approach, or separation in profile space; it does not establish collaboration, topical merger, or intellectual influence.

Chapter~9 compresses these findings into typologies and gives the final empirical contrast. Static morphology is weakly clustered. The selected Ward \(k=5\) typology has silhouette \(=0.133\), which supports descriptive anchors rather than natural categories. Those anchors are still useful: broad sparse profiles, compact low-dispersion profiles, low-dimensional locally uneven profiles, temporal novelty profiles, and uneven dispersion outliers give names to recurrent combinations of the eleven metrics. But the geometry remains continuous. The temporal trajectory typology is clearer and more stable: the selected dynamic solution has silhouette \(=0.396\) and mean subsample ARI \(=0.725\). Subfields are easier to group by how their morphology changes than by their full-period shape.

The centroid and hybrid sensitivity checks complete the interpretation. Clustering subfield centroid embeddings separates somewhat more clearly and aligns more strongly with OpenAlex domains than morphology clustering does. Hybrid clustering becomes highly stable when semantic location receives substantial weight. These results do not undermine the morphology typology; they show that semantic location and morphology answer different questions. A field's position in embedding space is not the same as the shape of its distribution around that position. The strongest final claim is therefore layered: static shape, temporal change, taxonomy, and semantic location provide different readings of scientific structure.

\section{Methodological Contributions}

The first methodological contribution is a reproducible corpus design for embedding-based science mapping. The thesis builds a synchronized OpenAlex title--abstract corpus covering 2000--2024, with annual balancing across subfield-year cells. This sampling strategy deliberately reduces the dominance of high-volume subfields and makes temporal and cross-sectional comparisons more defensible. It also makes the limits of the empirical object explicit: the corpus is English-only, restricted to articles and preprints, dependent on title and abstract availability, and organized by primary-topic subfield.

The second contribution is the disciplined use of SPECTER2 as an analytical geometry. The thesis computes quantitative metrics in the original embedding space rather than on UMAP, PCA, or other low-dimensional projections. This is a small technical choice with large interpretive consequences. It prevents visual maps from silently becoming measurement spaces, and it keeps the distinction between representation, measurement, and communication visible throughout the analysis.

The third contribution is the reduced eleven-metric morphology core. The metric set covers global dispersion, local density and hubness, spectral structure, and temporal movement. No single metric is asked to summarize a field. Instead, morphology is treated as a profile. This makes it possible to see, for example, why a field can be dispersed but locally regular, compact but hub-concentrated, or spectrally simple while temporally novel.

The fourth contribution is the separation of static morphology, temporal morphology, morphological similarity, and semantic-location clustering. These are often conflated in visual science maps because a single two-dimensional display can make position, distance, cluster, and category appear to be the same object. The thesis keeps them apart. Static profiles describe full-period shape. Temporal trajectories describe change in windowed structure. Similarity compares profiles between fields or subfields. Centroid clustering checks semantic location rather than shape.

The fifth contribution is interpretive conservatism in clustering. The thesis does not choose the most visually attractive partition or turn silhouette scores into taxonomy. Static clusters are treated as typological anchors because they are weak. Temporal clusters are treated as dynamic tendencies because they are clearer but still exploratory. Centroid and hybrid clusters are used as sensitivity evidence, not as replacements for morphology. This is an epistemological contribution as much as a technical one: it shows how to use clustering without pretending that every partition is a discovery of natural kinds.

\section{Limitations and Threats to Interpretation}

The main data limitations follow from the corpus design. OpenAlex is open and reproducible, but it is not a neutral census of science. Coverage differs across fields, years, languages, document types, and metadata practices. Primary-topic assignment forces each work into one subfield, even when papers have multiple intellectual affiliations. The English-only filter improves comparability but excludes other scientific language communities. The article and preprint restriction removes document types that may matter in some fields. The title--abstract requirement produces a cleaner semantic input while underrepresenting records with weak abstract coverage. The annual sampling cap and sparse subfield-year cells further mean that the corpus is a controlled analytical sample rather than the observed universe of publication.

The representation also imposes limits. SPECTER2 embeddings are learned document representations, not ground truth. They compress titles and abstracts into vectors shaped by training data, citation-informed supervision, and scientific-language coverage. They may encode patterns that are useful for scientific document retrieval while still missing method, evidence, theoretical stance, institutional context, or domain-specific nuance. A title--abstract embedding cannot represent the full structure of a paper, let alone the social organization of the field that produced it.

The geometric summaries carry their own risks. High-dimensional spaces are affected by anisotropy, distance concentration, local density variation, and hubness. Some of these phenomena are measured in the thesis, but they also shape the measurements themselves. The metrics are summaries of distributions, not full distributional comparisons. Centroid dispersion, kNN distances, hub Gini, and spectral proxies each compress many paper-level relations into a small number of statistics. PCA \(D_{80}\) is an operational proxy for spectral complexity, not a claim about true intrinsic dimension. PCA, PCoA, and UMAP projections are useful for communication, but they are not the spaces in which the quantitative claims are computed.

Clustering is the final limitation to keep in view. The static typology is weak by standard separation diagnostics and should not be treated as evidence that science has five morphological categories. The temporal typology is clearer and more stable, but it is still a descriptive partition of slope profiles, not a natural taxonomy of scientific change. Centroid and hybrid clustering show that semantic location can reconstruct more domain-like structure, but that is precisely why those results cannot be interpreted as morphology alone. The validity of the thesis does not depend on any cluster being real in an ontological sense; it depends on the usefulness of the measured distinctions and on the honesty of their scope.

\section{Implications and Future Research}

The framework suggests a broader implication for science mapping: fields can be compared by structural shape, not only by topical proximity or citation connectivity. This opens a descriptive diagnostic perspective. A field can be monitored for broadening, contraction, local density change, hub concentration, or spectral simplification without treating those movements as quality signals. The dynamic typologies in Chapter~9 are especially useful in this sense. They name patterns of movement--compacting, broadening, smoothing, and dimensionalizing--that could guide closer historical or domain-specific investigation.

Any practical use should remain cautious. Morphology should not be used to rank fields, allocate resources mechanically, or infer intellectual maturity. A compact field is not better than a dispersed one. A field with high novelty is not necessarily more innovative. A field whose morphology diverges from others is not necessarily isolated. The value of the framework is diagnostic: it can identify where structural changes or unusual profiles occur, so that substantive experts can ask better questions.

Several extensions follow directly from the thesis. One is representational comparison: the same morphology pipeline could be run with other scientific embedding models to test which findings are stable across representations and which are SPECTER2-specific. A second is linguistic expansion: multilingual corpora could show how much of the observed morphology is tied to English-language title--abstract records. A third is text-depth expansion: full-text or section-aware embeddings could separate abstract-level positioning from methods, results, and discussion structure.

The relational side could also be expanded. Morphology could be compared with citation networks, journal structures, funding categories, institutional affiliations, or author mobility to ask when shape, topic, and social organization move together. More complete distributional distances--for example Wasserstein, maximum mean discrepancy, or Gaussian/Frechet approximations--could compare subfield distributions beyond the reduced metric profile. Temporal analysis could move from five-year windows to annual models, change-point detection, or event-based histories. Finally, paper-level communities could be studied separately, provided the distinction remains clear: clustering papers by topic is not the same as measuring the morphology of fields.

The point of these extensions is not to replace the present framework with a larger map. It is to preserve the key distinction while adding evidence around it. Scientific fields have positions, shapes, trajectories, and relations. Measuring those dimensions separately makes science more legible, but also more resistant to any single, simple map.
