% Appendix C: Declaration on the Use of Generative Artificial Intelligence

\label{app:ai-declaration}

\noindent This appendix declares the limited use of generative artificial intelligence (generative AI) during the preparation of this thesis. OpenAI tools, specifically ChatGPT and Codex in the versions available during the 2025--2026 academic year, were used as auxiliary support. Their output was not treated as an academic source, and responsibility for the content, analysis, and conclusions remains entirely with the author.

\medskip
\noindent\textbf{Responsible use.}
No personal or confidential data were provided to generative AI systems. No copyrighted material was submitted beyond uses permitted for consultation, quotation, or research. The tools were used in accordance with their terms of use, and all suggestions were critically reviewed before they were incorporated. Generative AI was not used to produce experimental data, empirical results, scientific interpretations, or conclusions.

\medskip
\noindent\textbf{Technical use.}
The tools were used occasionally for two restricted purposes. First, they assisted with the detection of typographical errors and minor grammatical improvements in passages previously written by the author, particularly in English. Second, they provided support in diagnosing specific Python and LaTeX errors. Any resulting suggestion was reviewed, adapted to the project, and tested by the author before use.

\noindent Representative queries included requests such as \enquote{Review the grammar of this paragraph without changing its technical meaning} and \enquote{Help me identify the cause of this Python/LaTeX error}. Generative AI was not used to determine the structure of the thesis, replace the literature review, select the methodology, or formulate its arguments and conclusions.

\medskip
\noindent\textbf{Assessment of usefulness and limitations.}
This limited use was helpful for identifying formal errors and accelerating mechanical revision and debugging tasks. Its main benefit was the immediate provision of suggestions that could subsequently be checked. However, the responses could be imprecise, alter the intended meaning, or propose code that did not fit the project context. For this reason, all suggestions were verified against the relevant sources, the empirical results, and the execution of the code.

\medskip
\noindent\textbf{Author:} Alejandro Treny Ortega

\noindent\textbf{Date:} 13 June 2026
